What We Eval in the Shadows

### 2021-08-09 paper/main.tex
Move subsections to Discussion section
Fix code macro and improve presentation of Simplify Interfaces section

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}

% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother

@@

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}

\subsection{Simplify Interfaces}
Many packages use the capability to parse and evaluate expressions in custom
environments to design simpler interfaces. For example, the \c{AdapSamp} package
provides functions that implement sampling algorithms. These functions take a
\c{formula} argument, an expression in \c{x} as a string, that represents the
kernel of the target distribution. This is turned into a closure using
\c{function(x){eval(parse(text=formula))}}. While the \c{formula} supplied by
the user could be a higher-order function, passing \c{"exp(-x\^2/2)"} as an
argument is simpler for domain experts compared to \c{function(x) exp(-x\^2/2)}.
The \c{bain::constraint\_to\_row} function has a similar pattern of \eval usage.
It takes a constraint equation as a string argument, and, parses and evaluates
it in a custom environment with the expected bindings. The
\c{ctsem::ctModelTransformsToNum} function similarly evaluates a vector of
formulae represented as a string vector.

\subsection{Custom Evaluation}
The \eval function is commonly used to customize the evaluation of expressions.
An example is the \c{assertthat} package's \c{see\_if} function which is used to
assert that the sequence of conditions supplied as arguments are true. It
evaluates the condition expressions one-by-one inside a \c{tryCatch} block and
produces an error at the first expression that returns \c{FALSE}. If the
expression itself produces an error, the error message is captured by the
enclosing \c{tryCatch} block and returned. The remaining expressions in the
sequence remain unevaluated. A similar pattern is used by the \c{checkmate}
package.

@@

The ability to capture the current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes R the toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
  defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
  subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

Another example is the \c{tidyselect} package that defines a DSL for selecting
columns in a data frame powering number of data manipulation operation in the
\c{data.table} main competitor, the \c{dply} package. It uses eval to build an
interpreter for evaluating column selectors such as \c{select(df,
starts\_with("length") \& !(name:mass))} which returns all columns in \c{df}
that starts with the string \c{length} and is not between \c{name} and
\c{mass} columns.

Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}, on expressions built with \c{expression()} or \c{parse}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created: it contains expressions that are derived twice (with \c{DD}). To get a concrete value, it is evaluated in  dataframe \c{mu} where symbolic variables are bound to actual numbers.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

The \c{glue} package provides functions for string interpolation; text enclosed
by c{\{} and c{\}} is evaluated using \eval as an R expression and replaced by
its result. The code snippet below shows an example of this.

@@

\subsection{Miscellaneous}
A common use of \eval is the extraction of unevaluated text of arguments passed
to \c{...} (vararg). For a non-vararg parameter \c{x}, the expressions
\c{substitute(x)} can be used to extract the argument text. However, for \c{...}
arguments, the expression \c{eval(quote(substitute(list(...))))} or its variants
are used. The \eval call evaluates the quoted expression
\c{substitute(list(...))}. This returns a \c{list} of unevaluated argument
expressions. The \c{hasGroupNumber} function from the \c{lattice} package is one
of the many users of this pattern.

The \c{withr} package provides the \c{defer} function which defers the
evaluation of an expression until a stack frame is exited. It attaches this
expression as an attribute to the frame environment and uses \eval to evaluate
it on frame exit.

@@

\Eval in R is used chiefly for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as a
function or method call, or an assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\c{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \c{get}, and assignment in any environment,
with \c{assign}. Building a call and executing can be done with \c{do.call}.
Nevertheless, many of the observed \eval usage in R, particularly, the ones
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

Finally, let us mention an anecdotal use of \eval as a means to bypass some
checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example,  it checks that packages
do not call the \c{unlockBinding} function.\footnote{It allows one to open and
modify loaded package namespaces.} Packages such as \c{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

@@

Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
\eval can be replaced with alternate functions yielding the same results. For
example, the \c{PerformanceAnalytics} package contains a function
\c{chart.QQPlot} that uses \eval to resolve a string into function and another
to call it and assign its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
  or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.

  The \c{copyslots} function from the \c{coin} package uses \eval to copy the
  slots from one \c{S4} object to another. It performs
  \c{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
  \c{s}. This can be replaced by the expression \c{slot(target, s) <- slot(source,
    s)} which uses the \c{slot<-} and \c{slot} functions to assign and retrieve
  the slots by name.

  Another examples is in the \c{configr::config.funs.par} function whose
  simplified definition is shown below.

@@

  The function appears to use \eval since the function name \c{fun} is received
  as a string. This string can be resolved to the function object using the
  \c{get} function and the contents of \c{text.1}, \c{text.2}, and \c{text.3}
  can be written directly without the need to convert them to text for
  evaluation using \eval.

>>>>

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\newcommand{\code}[1]{\lstinline |#1|\xspace}
%\renewcommand{\c}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother

@@

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}
To understand the how R developers use \eval, we manually inspected top 117 eval
call-sites which contribute to XX\% of the eval calls. We describe the common
usage patterns below.

\subsection{Simplify Interfaces}

A common use of \eval is to \emph{simplify interfaces} by letting users pass string
argument which can be parsed and evaluated in custom a environment. An example
is the \code{rARS} function from the \code{AdapSamp} package, shown below.

\begin{lstlisting}
rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
  ...
  p <- function(x) { eval(parse(text=formula)) }
  ...
}

\end{lstlisting}
The \code{rARS} function take a \code{formula} argument, an expression in
\code{x} as a string, that represents the kernel of the target distribution.
This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
  \}}. An alternative design would be to supply the closure itself as the
argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
convenient for domain-experts as passing \code{"exp(-x^2/2)"}.

\subsection{Custom Evaluation}
The \eval function is commonly used to customize the evaluation of expressions.
An example is the \code{assertthat} package's \code{see\_if} function which is used to
assert that the sequence of conditions supplied as arguments are true. It
evaluates the condition expressions one-by-one inside a \code{tryCatch} block and
produces an error at the first expression that returns \code{FALSE}. If the
expression itself produces an error, the error message is captured by the
enclosing \code{tryCatch} block and returned. The remaining expressions in the
sequence remain unevaluated. A similar pattern is used by the \code{checkmate}
package.

@@

The ability to capture the current call as a language object (\code{match.call()}),
extract the unevaluated expression from function arguments (\code{substitute()})
and \eval with first class environment makes R the toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \code{data.table}
package defines a DSL for data frame manipulation using the subset operator
\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \code{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

Another example is the \code{tidyselect} package that defines a DSL for selecting
columns in a data frame powering number of data manipulation operation in the
\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
interpreter for evaluating column selectors such as \code{select(df,
starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
that starts with the string \code{length} and is not between \code{name} and
\code{mass} columns.

Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}. They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}). To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

The \code{glue} package provides functions for string interpolation; text enclosed
by c{\{} and c{\}} is evaluated using \eval as an R expression and replaced by
its result. The code snippet below shows an example of this.

@@

\subsection{Miscellaneous}
A common use of \eval is the extraction of unevaluated text of arguments passed
to \code{...} (vararg). For a non-vararg parameter \code{x}, the expressions
\code{substitute(x)} can be used to extract the argument text. However, for \code{...}
arguments, the expression \code{eval(quote(substitute(list(...))))} or its variants
are used. The \eval call evaluates the quoted expression
\code{substitute(list(...))}. This returns a \code{list} of unevaluated argument
expressions. The \code{hasGroupNumber} function from the \code{lattice} package is one
of the many users of this pattern.

The \code{withr} package provides the \code{defer} function which defers the
evaluation of an expression until a stack frame is exited. It attaches this
expression as an attribute to the frame environment and uses \eval to evaluate
it on frame exit.

@@

\Eval in R is used chiefly for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as a
function or method call, or an assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\code{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \code{get}, and assignment in any environment,
with \code{assign}. Building a call and executing can be done with \code{do.call}.
Nevertheless, many of the observed \eval usage in R, particularly, the ones
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

Finally, let us mention an anecdotal use of \eval as a means to bypass some
checks performed by the \code{R CMD CHECK} tool. This tool performs a number of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example,  it checks that packages
do not call the \code{unlockBinding} function.\footnote{It allows one to open and
modify loaded package namespaces.} Packages such as \code{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

\section{Discussion}

@@

Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
\eval can be replaced with alternate functions yielding the same results. For
example, the \code{PerformanceAnalytics} package contains a function
\code{chart.QQPlot} that uses \eval to resolve a string into function and another
to call it and assign its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
  or even to a one-liner \code{do.call(paste0("q",d), c(dp, list(...)))}.

  The \code{copyslots} function from the \code{coin} package uses \eval to copy the
  slots from one \code{S4} object to another. It performs
  \code{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
  \code{s}. This can be replaced by the expression \code{slot(target, s) <- slot(source,
    s)} which uses the \code{slot<-} and \code{slot} functions to assign and retrieve
  the slots by name.

  Another examples is in the \code{configr::config.funs.par} function whose
  simplified definition is shown below.

@@

  The function appears to use \eval since the function name \code{fun} is received
  as a string. This string can be resolved to the function object using the
  \code{get} function and the contents of \code{text.1}, \code{text.2}, and \code{text.3}
  can be written directly without the need to convert them to text for
  evaluation using \eval.
